A Guide to Designing Experiments to Test Statistical Graphics

### 2025-03-14 submitted-index.tex
Review modifications

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\usepackage{bookmark}

@@

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2025-03-14}

@@

Data visualizations are a critically important tool for communicating
scientific information to the public in what creators hope is an
easy-to-digest, visually attractive form. There are many strategies for
creating charts and graphs, from Tufte-esque minimalism (Tufte, 1991) to
charts designed with extra imagery and aesthetic appeal that draw the
viewer's attention and persist in memory (Cairo, 2012). For a specific
type of data, there are also usually many different chart forms to
display that data: for instance, if we have a set of categorical data
and we wish to show the relative proportions of each category, we could
do so using a stacked bar chart or the polar equivalent, a pie chart.
There have been several attempts to list out all of the types of charts
(Ribecca, 2022), create a taxonomy of charts (Bertin \& Berg, 1983;
Desnoyers, 2011), and even to create charts using a domain-specific
grammar of graphics (Wilkinson, 1999) that is also useful for
classification. One extremely useful reference is from Data to Viz
(https://www.data-to-viz.com/), which uses a decision tree to show
different visualizations compatible with the data; R, python, D3.js, and
React code are provided to demonstrate how to create those
visualizations. With all of the different design choices available, how
are chart creators to know what is the best approach for communicating
data to the appropriate audience?

While there are heuristics, rules-of-thumb, and guidelines (Allen \&
Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on Standards
for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating useful
and visually attractive data displays, the best way to establish the
efficacy of various design decisions is to test the visualization on
humans, evaluating different variants under controlled conditions
(Cleveland et al., 1988; Cleveland \& McGill, 1985). Empirical
assessments of visualizations, when carefully designed, allow
statisticians to determine which representation of the same data is most
effective along one or more dimension(s) of interest: estimation or
prediction accuracy, within or between group comparisons, response time,
and ability to draw real-world decisions are common goals for charts.

It is extremely challenging to design studies which strike the right
balance between experimental control (i.e.~internal validity) and
generalizability to a wider context (i.e.~external validity). Simply
asking people to read quantities off of a graph may not generalize
beyond the questions asked or the data used in the chart (Croxton, 1932;
Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but designing a
study that is sufficiently robust to those issues requires manipulation
or control of so many factors that the amount of participants and trials
quickly becomes daunting or unaffordable. In this paper, we attempt to
distill the experience gained from conducting several different types of
graphics experiments (Hofmann et al., 2012; Robinson, 2022; VanderPlas
et al., 2019; Vanderplas et al., 2024; VanderPlas \& Hofmann, 2015,
2017), discussing the use of different testing methods (Vanderplas et
al., 2020), the process of designing a graphical experiment, and
analysis of the resulting empirical data. It is our hope that this paper
will lower the barriers that exist for conducting empirical graphics
research and reduce the probability of costly mistakes.

@@

There are many different testing methods used to empirically assess
statistical graphics. This paper uses studies conducted online without
additional equipment as primary examples, though many of the same
considerations apply to in-person experiments conducted using additional
equipment, including 3D printed charts, eye-tracking equipment, and
interactive data displays. Online experiments have lower overhead, offer
relatively fast data collection, and provide useful results for
well-designed experiments. The toolkit used for these experiments is
R-based (R Core Team, 2022), and includes ggplot2 (Wickham, 2016) and
Shiny (Chang et al., 2021) as primary components. In many experiments,
we customized the Shiny interface with JavaScript and d3 (Bostock et
al., 2011), enabling interactive graphics, use of svgs, and other useful
extensions. While we prefer this set of tools, most of the observations
described here apply to a wide variety of different workflows for
graphical experimentation, including in-person experiments.

It is important to consider the level of user engagement which is
necessary to complete a particular visual or graphical task. For
instance, testing whether someone can detect an effect such as a linear
trend in noisy data is a perceptual question. Perceptual questions are
often examined experimentally using methods which allow the user to
interact with the data on a basic visual level: users are presented with
a visual stimulus and answer yes/no questions to indicate whether the
effect is detected. Numerical estimation is another common task when
testing graphics: in these experiments, the participant views a chart,
estimates the requested numerical quantity, and enters the estimate into
the application through a numerical input, slider, or other form
element. Sometimes, it is possible to set up a scenario where the user
adjusts the plot using a set of controls designed to provide a fixed set
of interactive operations. This type of user engagement was used to
assess the strength of the sine illusion (VanderPlas \& Hofmann, 2015):
users adjusted the strength of a transformation designed to correct the
illusion until the lines appeared to be the same length, as shown in
Figure~\ref{fig-sine-illusion}, providing a direct measure of the
magnitude of the sine illusion's effect. In other situations, it may be
preferable to have the user directly interact with the visual stimulus.
In Vanderplas et al. (2024), participants were asked to rotate and
interact with a 3D rendered bar chart; the application recorded user
interactions and corresponding rotation matrices, providing insight into
the visual comparisons the user may have been performing. This
information was used as a supplement to the explicitly provided
estimates, providing some contextual information as well as the ability
to identify the level of participant engagement with the questions. When
experiments are conducted as part of classroom experiential learning, it
is sometimes helpful to be able to separate the low-effort participants
from those who were fully intellectually engaged in the task.
Interactive graphics provide another level of user engagement that can
be much more open-ended. With interactive graphics, researchers can ask
participants to directly annotate plots, toggle aesthetics, and
highlight groups and plot features. Careful implementation of the
experiment application may allow for each of these interactions to be
recorded and analyzed, producing a rich, if messy, set of data that may
allow researchers to tease apart visual estimation error from common
shortcuts such as rounding used during direct numerical estimation.

@@

\pandocbounded{\includegraphics[keepaspectratio]{sine_illusion_screenshot.png}}

@@

\caption{\label{fig-sine-illusion}Direct adjustment of a plot in a
perceptual task. In this experiment, designed to assess the strength of
the sine illusion, the user adjusts the plot using - and + buttons,
which control the strength of a transformation designed to correct the
effect of the sine illusion. When the user is satisfied that the lines
are of equal length, they select the `Finished' button to move to the
next task. The experiment used a psychophysics experimental design, the
method of adjustment, but leveraged the interactive Shiny interface to
record the entire sequence of adjustments made by the user for each
trial. A demo version of this application can be found at
https://shiny.srvanderplas.com/sine-illusion/.}

@@

Visual inference (Buja et al., 2009; Wickham et al., 2010) is another
useful testing tool for perceptual questions such as ``which chart
displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the statistical significance of the graphical
finding in a chart. Visual inference charts are often called ``lineups''
in analogy to the criminal procedure where the suspect is placed in a
line with several other individuals with similar characteristics. In a
graphical lineup procedure, there is a target plot containing the real
data, embedded in an array of innocent ``null'' plots generated through
resampling or simulation. If viewers consistently pick the target plot
at a higher rate than any of the null plots, the target plot is said to
be visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013)
and a ``see'' value, the visual analogue of a \(p\)-value (Chowdhury et
al., 2020), can be calculated using the \texttt{vinference} R package or
the process described in VanderPlas et al. (2021). The details of this
calculation are beyond the scope of this broader discussion of how to
test charts, but more detail on visual inference is provided in
\textless insert citation to visual inference WIRE article under
development\textgreater.

In another variation of the statistical lineup procedure, data generated
from two models are compared, with target plots from each model embedded
in the array of \(K\) total plots. The \(K-2\) null plots are
constructed from a mixture model that blends the two competing models
(VanderPlas \& Hofmann, 2017). Viewers are asked to select the panel(s)
which are most different, and the primary source of information are
trials in which viewers selected the target from one model but not the
other, indicating that the display method used allowed viewers to
differentiate one model's data (but not the other) from the nulls
created through a mixture model. This variation allows the experimenter
to assess graphical design choices to determine whether they effectively
emphasize structural differences in the data (VanderPlas \& Hofmann,
2017).

One advantage of the visual inference technique is that the experimenter
can ask a very general question, such as ``which of these plots is the
most different?'', rather than a specific question about the displayed
data which may require more quantitative sophistication. All of the
necessary information to make the decision is embedded in the choice of
the model used to generate the null plots. This feature is extremely
convenient when conducting the experiment and even allows small children
to complete the task. The downside is that as a result, visual inference
experiments do not allow experimenters to assess the viewer's
understanding of the information shown in the chart. In most cases,
visual inference experiments remove any contextual information from the
charts, including axis labels and values, plot titles, and so on, in
order to encourage participants to make decisions based solely on the
graphical presentation. This lack of context is a double-edged sword:
visual inference can involve participants who do not have any
mathematical training or instincts (including children), but researchers
also cannot use this technique to assess higher levels of engagement
with a chart, such as estimation, prediction, or reasoning based on
displayed information.\\

To assess the viewer's \emph{understanding} of information shown in a
chart, we must ask questions and allow the user to provide feedback.
User feedback may be collected on a numerical scale or through the use
of written comments, recorded ``think-aloud'' processes, and other more
qualitative interaction methods. In some studies, asking users to
interpret a chart within a larger scenario can be effective, as in
Figure~\ref{fig-estimation-describe}, while in others it is more helpful
to ask users to explain answers. In visual inference studies, asking
users why a specific panel was chosen has been demonstrated to provide
rich insight into otherwise confusing numerical results (VanderPlas \&
Hofmann, 2017).

Think-aloud methods ask the viewer to narrate their internal thought
process, either during or after completing a task (Haak et al., 2003).
These recordings (or transcripts) can provide valuable insights into
conscious cognition, and are often used when conducting usability
studies. While we have not to date recorded users talking out loud about
what they are seeing during a study, think aloud methods could easily be
implemented within a Shiny application, with audio recordings saved to
the server for transcription and analysis (Dunbar, 1995; Kirschenbaum,
2003; Trafton et al., 2000). It is even possible that these recordings
could be automatically transcribed using speech-to-text models. We have
used think-aloud methods informally during pilot studies to ``harden''
graphical experiments and verify the selection of parameters used in an
experiment. The success of this approach, combined with the few studies
which used think-aloud to assess charts (Haider et al., 2021; Kulhavy et
al., 1992; Lee et al., 2016), suggests that think-aloud methods are an
often-overlooked but useful tool for assessing data visualizations.

@@

\pandocbounded{\includegraphics[keepaspectratio]{Estimation_describe_plot_crop.png}}

@@

Of course, in an online, asynchronous experiment, every user interaction
with the testing materials (typically hosted on a web page) can also be
recorded along with time stamps, mouse positions, browser size and
screen resolution, and other information. While we have not used this
type of information heavily in our experimental analyses thus far, in
most experiments we collect time stamp data in order to assess how long
participants spend on each question. Typically, the first round of test
questions takes the longest for participants to complete. Additional
replicates do not usually affect accuracy (i.e.~there is no immediate
learning effect) until after `too many' tests cognitive fatigue proves
to be detrimental to accuracy (Chowdhury et al., 2018). This sweet spot
between replicates and fatigue depends on the cognitive burden in each
test and should factor into designing the experiment. In some
experiments, we have provided participants with supportive tools, such
as ``scratch pads'' and calculators built into the Shiny application to
support the complex calculations required to answer higher-level
numerical estimation questions (Figure~\ref{fig-estimation-calc}). In
order to be supportive, the tools must be easy to use, but assuming this
bar is met, the tools can reduce participant cognitive load while
recording a wealth of information. This information provides real
insight into how participants were looking at the data, what strategies
they tried and discarded for reading the chart, and what visual
estimation methods were used. While systematic analysis and modeling of
this data may be difficult, as it is usually messy and often must be
manually coded, the insights provided can be extremely useful. However,
unless participants are required to use these tools, it is difficult to
gather comprehensive information - those participants that don't use
supportive tools likely differ in meaningful ways from those who do. As
a result, the information gathered from supportive tools likely does not
generalize to the entire sample.

@@

\pandocbounded{\includegraphics[keepaspectratio]{Estimation_numerical_screenshot_crop.png}}

@@

The choice of the parameter space used in testing is crucial to gain
insight from a study without putting too much burden on participants
with overlong studies. Choosing an appropriate space for testing
parameters is a well-known problem in psychometric testing: the space
considered should cover the area between `only some activation' to
`almost full activation' of an appropriate psychometric function (Schütt
et al., 2016; Valentin et al., 2024). When testing charts, visual
assessment is obviously key, but researchers can make use of statistical
indices related to the testing condition to narrow the parameter space
to a reasonable and efficient subset from which maximal information can
be acquired.\\
These statistical indices may also serve as quantitative proxies for the
difficulty of the visual task.

To identify a statistical proxy for visual difficulty that may help with
narrowing the parameter space, it can be useful to consider numerical
measures used to estimate the same types of visual information that will
be assessed in the experiment. For instance, we have used

@@

\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-lof-density-curves-1.pdf}}

@@

While this method is certainly more critical for model-based sampling
methods, it is also important when data are generated by sampling from a
larger data set. When sampling from a larger dataset, parameters are
more often sample size and stratification methods, but it is still
important to iteratively assess the data generating procedure through
simulation. Using numerical proxies for visual characteristics of data
displays such as curvature, linearity, scatter, dispersion can assist
with identifying optimal parameter settings to use across different
experimental conditions. Even with this strategy, it is still critical
to fine-tune the parameter choices with visual calibration and pilot
testing.

@@

Once an appropriate set of parameters are identified using the numerical
screening method, it is important to calibrate these parameter
selections visually. No numerical statistic is a perfect measure of what
we actually see: at best, they are approximations of what we might
potentially see. We have found it to be useful to have one experimenter
calibrate the model parameters at a gross level, and then have another
experimenter narrow in on the parameters which are visually reasonable
within the selected range. Then, both examiners visually inspect a large
number of plots generated using those parameters to get a sense for how
difficult the task at hand is (this strategy is also described by Lu et
al. (2022)). At some point, all experimenters become so visually
saturated with the nuances of the data generating mechanism that it may
become necessary to ``sanity check'' the protocol with family members,
friends, and colleagues. These informal surveys provide extremely useful
feedback and can help to counteract the visual saturation of being
immersed in the design of a visualization experiment for months at a
time.

@@

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing before deploying an
experiment. In support of this assessment, we offer up a cautionary tale
of our own experience: that of VanderPlas \& Hofmann (2017), where we
designed an experiment to test which plot aesthetics promoted discovery
of linear trends and/or clusters.

@@

\pandocbounded{\includegraphics[keepaspectratio]{lineup-missing-ellipse.png}}

@@

\caption{\label{fig-lineup-problems}A lineup from Vanderplas \& Hofmann
(2017). Panel 10 shows the clustered target data and panel 17 shows the
target data with a strong linear relationship; either of these target
panels was the expected choice. Unfortunately, panel 16 has only two
bounding ellipses shown, which is an unintentional difference that
resulted from a faulty method for assigning clusters to null plots; many
participants selected this panel instead of one of the target panels.}

@@

One reason why it is so difficult to generate sampling models for visual
explorations is that our visual system is optimized for identifying
differences between groups. This ability can interfere with the natural
to use the null sampling models that might be used in equivalent
numerical tests when running experiments that use visualizations. We
re-ran the experiment using a different clustering method that
controlled the number of points in each group. Instead of noticing the
number of ellipses, participants instead used the differences in size
and shape of the ellipses formed when clustering after the data
generating procedure. That is, participants could still detect the
artificial nature of the induced clusters using other features. While it
can be difficult to get the data generating method right, it is
essential to conducting visual experiments that generalize well beyond
the effects shown in a single data set or phenomenon. The time and
effort invested in this step at the outset of the experiment pays
dividends when it allows for clear generalization of the experimental
results to an entire statistical concept rather than a single data set.

@@

\subsection{Cognitive Load}\label{cognitive-load}

@@

\subsection{Participant Instructions}\label{participant-instructions}

@@

\subsection{Accounting for Participant
Variability}\label{accounting-for-participant-variability}

@@

\subsection{Demographic Data}\label{demographic-data}

It may be useful to ask a few more demographic questions about STEM
education level for studies which ask more of participants from a
mathematical standpoint; while lineup studies have not found strong
associations with those variables, lineup studies also do not require
participants to engage with the data presented in a chart in a way that
requires higher-order mathematical reasoning. This has allowed us to
make the argument to the ethics committee (institutional review board,
or IRB) that our research is exempt, as we do not collect enough
demographic information to identify participants, however, collecting
reduced demographic information occasionally comes at a cost. One recent
study examining the use of log scales and exponential data was conducted
using Prolific, which recruits participants from around the world; we
required only that participants were fluent in English to participate.
It was only after the experiment was completed that we realized that
different countries introduce logarithms as a concept at different
points during primary and secondary education; it might be that
individuals in some countries have much more experience with log scales
than those educated in the United States. We mention this only to point
out that while every experiment contains a few missed opportunities, it
is worth giving careful thought to the demographic questions asked of
participants and what information may be helpful during the analysis
stage.

\subsection{Trial Allocation to
Participants}\label{trial-allocation-to-participants}

@@

\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-density-rug-1.pdf}}

@@

In this paper, we have attempted to contextualize and motivate the logic
behind the process we use to design empirical graphics experiments.
We've discussed model development, experimental design considerations,
pilot testing, and data analysis methods that have been honed over many
successful and less-than-optimal experiments. While no experiment
involving humans ever goes exactly to plan, following this process helps
to avoid some of the most likely mishaps, ensuring that each new
experiment's ``bonus'' findings have only minimal impacts on the study's
overall utility.

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\title{A Guide to Designing Experiments to Test Statistical Graphics}
\author{Emily Robinson \and Heike Hofmann \and Susan Vanderplas}
\date{2024-11-01}

@@

Charts and data visualizations are primary methods of communicating
scientific information to the public in what creators hope is an
easy-to-digest, visually attractive form. There are many strategies for
creating charts and graphs, from Tufte-esque minimalism (Tufte, 1991) to
charts designed with extra imagery and aesthetic appeal that draw the
viewer's attention and persist in memory (Cairo, 2012). For a specific
type of data, there are also usually many different chart forms to
display that data: for instance, if we have a set of categorical data
and we wish to show the relative proportions of each category, we could
do so using a stacked bar chart or the polar equivalent, a pie chart.
There have been several attempts to list out all of the types of charts
(Ribecca, 2022), create a taxonomy of charts (Bertin \& Berg, 1983;
Desnoyers, 2011), and even to classify charts using a domain-specific
grammar of graphics (Wilkinson, 1999). With all of the different design
choices available, how are chart creators to know what is the best
approach for communicating data to the appropriate audience?

While there are heuristics, rules-of-thumb, and various guidelines
(Allen \& Erhardt, 2016; Few, 2006; Haemer, 1948; {``Joint Committee on
Standards for Graphic Presentation,''} 1915; Kosslyn, 2006) for creating
useful and visually attractive charts, the best way to establish the
efficacy of various design decisions is to test charts: evaluate them
under controlled conditions, and gather appropriate data (Cleveland et
al., 1988; Cleveland \& McGill, 1985) to compare different
representations of the same data. It is extremely challenging to design
studies which strike the right combination of internal and external
validity. Simply asking people to read quantities off of a graph may not
generalize beyond the questions asked or the data used in the chart
(Croxton, 1932; Croxton \& Stryker, 1927; Eells, 1926; Huhn, 1927), but
designing a study that is sufficiently robust to those issues requires
manipulation of so many factors as to be functionally close to
impossible to control or manipulate in a single experiment. In this
paper, we attempt to distill the experience gained from conducting
several different types of graphics experiments (Hofmann et al., 2012;
Robinson, 2022; VanderPlas et al., 2019; Vanderplas et al., 2024;
VanderPlas \& Hofmann, 2015, 2017), discussing the use of different
testing methods (Vanderplas et al., 2020), the process of designing a
graphical experiment, and analysis of the resulting empirical data. It
is our hope that this paper will lower the barriers that exist for
conducting empirical graphics research and reduce the probability of
costly mistakes.

@@

There are many different testing methods used to empirically assess
graphics. While it is certainly possible to use equipment such as
eye-tracking devices, this paper focuses primarily focus on studies
conducted without additional equipment, and primarily online, because
these experiments have the lowest overhead, offer relatively fast data
collection, and provide useful results, for well-designed experiments.
The toolkit used for these experiments is R-based (R Core Team, 2022),
and includes ggplot2 (Wickham, 2016) and Shiny (Chang et al., 2021) as
primary components, as well as JavaScript (including D3 (Bostock et al.,
2011)) to customize the Shiny interface for many of the experiments.
While we prefer this set of tools, most of the observations described
here apply to a wide variety of different workflows for graphical
experimentation.

One primary concern when testing graphics is the level of user
engagement required. For instance, testing whether someone can detect an
effect such as a trend line in noisy data is a \emph{perceptual}
question best addressed using methods which allow the user to interact
with the data on a basic visual level. Sometimes, it is possible to set
up a scenario where the user adjusts the plot directly, as in an
experiment testing the strength of the sine illusion experiment
(VanderPlas \& Hofmann, 2015): users adjusted the strength of the
illusion correction until the lines appeared to be the same length,
providing a direct measure of the magnitude of the sine illusion's
effect, as in Figure~\ref{fig-sine-illusion}. In other situations, it
may be possible to have the user directly interact with the plot, as in
Vanderplas et al. (2024), where participants were asked to rotate and
interact with a 3D rendered bar chart; the application then records all
relevant user interactions to provide insight into the visual
comparisons the user may have been performing as a supplement to the
explicitly provided estimates.

@@

\includegraphics{sine_illusion_screenshot.png}

@@

\caption{\label{fig-sine-illusion}Direct adjustment of a plot in a
perceptual task}

@@

Statistical lineups (Buja et al., 2009; Wickham et al., 2010) are
another useful testing tool for perceptual questions such as ``which
chart displays this data more clearly'' (Hofmann et al., 2012) while
simultaneously assessing the significance of the graphical finding in
the chart. As in the criminal procedure, lineups embed a target plot
with signal into an array of 19 other innocent ``null'' plots that show
only noise. If viewers consistently pick the target plot at a higher
rate than any of the 19 null plots, the target plot is said to be
visually significant (Loy \& Hofmann, 2013; Majumder et al., 2013) and a
``see'' value (the visual analogue of a \(p\)-value for a statistical
test) can be calculated (Chowdhury et al., 2020; VanderPlas et al.,
2021). In another variation of the statistical lineup procedure, two
models are compared, with target plots from each model embedded in the
array of 20 total plots; the remaining null plots are constructed from a
mixture model that blends the two competing models. This variation
allows the experimenter to assess graphical design choices to determine
whether they effectively emphasize structural differences in the data
and assign a hierarchy of visual attributes (VanderPlas \& Hofmann,
2017). Statistical lineups do not require the experimenter to provide
any contextual information for the data: all of the necessary
information is embedded in the choice of a null statistical model, which
is convenient for testing purposes, but does not allow for testing the
viewer's understanding of the information shown in the chart.

To assess the viewer's \emph{understanding} of information shown in a
chart, we must ask questions and allow the user to provide feedback.
User feedback may be collected on a numerical scale or through the use
of written comments, recorded ``think-aloud'' processes, and other more
qualitative interaction methods. In some studies, asking users to
interpret a chart within a larger scenario can be effective, as in
Figure~\ref{fig-estimation-describe}, while in others it is more helpful
to ask users to explain answers. In lineup studies, asking users why a
specific panel was chosen can provide rich insight into otherwise
confusing numerical results. While we have not (to date) recorded users
talking out loud about what they are seeing, think aloud methods could
be implemented within a Shiny application, with audio recordings saved
to the server for transcription and analysis (Dunbar, 1995;
Kirschenbaum, 2003; Trafton et al., 2000); it is even possible that
these recordings could be automatically transcribed using text-to-speech
functions of large language models.

@@

\includegraphics{Estimation_describe_plot_crop.png}

@@

Of course, in an online, asynchronous experiment, every user interaction
with the testing materials (typically hosted on a web page) can also be
recorded along with time stamps, mouse positions, browser size and
screen resolution, and other information. While we have not used this
type of information heavily in our experimental analyses thus far, in
most experiments, we collect time stamp data in order to assess how long
participants spend on each question. Typically, the first round of test
questions takes the longest for participants to complete. Additional
replicates do not usually affect accuracy (i.e.~there is no immediate
learning effect) until after `too many' tests cognitive fatigue proves
to be detrimental to accuracy (Chowdhury et al., 2018). This sweet spot
between replicates and fatigue depends on the cognitive burden in each
test and should factor into designing the experiment. In some
experiments, we have provided participants with supportive tools, such
as ``scratch pads'' and calculators built into the Shiny application to
support the complex calculations required to answer higher-level
numerical estimation questions (Figure~\ref{fig-estimation-calc}). In
order to be supportive, the tools must be easy to use, but assuming this
bar is met, the tools can reduce participant cognitive load while
recording a wealth of information. This information provides real
insight into how participants were looking at the data, what strategies
they tried and discarded for reading the chart, and what visual
estimation methods were used. While systematic analysis and modeling of
this messy data may be difficult, the insights provided can be extremely
useful.

@@

\includegraphics{Estimation_numerical_screenshot_crop.png}

@@

The choice of the tested space is crucial to gain insight from a study
without putting too much burden on participants with overlong studies.
Choosing an appropriate space for testing parameters is a well-known
problem in psychometric testing (\textbf{shah?}): the space considered
should cover the area between `only some activation' to `almost full
activation' of an appropriate psychometric function. When testing
charts, visual assessment is obviously key, but researchers can make use
of statistical indices related to the testing condition to narrow the
parameter space.\\
These indices then serve as quantitative proxies of visual difficulty.

If using model based sampling methods, it is important to iteratively
assess the sampling procedure through simulation. Start with a sampling
method and define a parameter space that is appropriate for the model,
and then simulate many data sets from each possible parameter
combinations using a grid search. When sampling from a large data set,
it is still important to assess the sample size and any potential
stratification methods employed in order to ensure that the visual
effect is represented.

In both cases, it can help to employ some general feature engineering -
what numerical statistic provides an estimation of the basic concept
which is being visually evaluated? For instance, we have used

@@

\includegraphics{index_files/figure-pdf/fig-lof-density-curves-1.pdf}

@@

@@

Once an appropriate set of parameters are identified using the numerical
screening method, it is important to calibrate these parameter
selections visually. No numerical statistic is a perfect measure of what
we actually see: at best, they are approximations of what we might
potentially see. We have found it to be useful to have one experimenter
calibrate the model parameters at a gross level, and then have another
experimenter narrow in on the parameters which are visually reasonable
within the selected range. Then, both examiners visually inspect a large
number of plots generated using those parameters to get a sense for how
difficult the task at hand is (this strategy is also described by Lu et
al. (Lu et al., 2022)). At some point, all experimenters become so
visually saturated with the nuances of the data generating mechanism
that it may become necessary to ``sanity check'' the protocol with
family members, friends, and colleagues. These informal surveys provide
extremely useful feedback and can help to counteract the visual
saturation of being immersed in the design of a visualization experiment
for months at a time.

@@

We cannot overstate the importance of visual assessment of your model
stimuli, preferably with fresh eyes. We highly recommend performing
several rounds of think-aloud pilot testing before deploying an
experiment. In support of this assessment, we offer up a cautionary tale
of our own experience: that of (VanderPlas \& Hofmann, 2017), where we
designed an experiment to test which plot aesthetics promoted discovery
of linear trends and/or clusters.

@@

\includegraphics{lineup-missing-ellipse.png}

@@

\caption{\label{fig-lineup-problems}A lineup from Vanderplas \& Hofmann
(VanderPlas \& Hofmann, 2017). Panel 10 shows the clustered target data
and panel 17 shows the target data with a strong linear relationship;
either of these target panels was the expected choice. Unfortunately,
panel 16 has only two bounding ellipses shown, which is an unintentional
difference that resulted from a faulty method for assigning clusters to
null plots; many participants selected this panel instead of one of the
target panels.}

@@

One reason why it is so difficult to generate sampling models for visual
explorations is that our visual system is very, very good at finding
differences and unexpected results. We re-ran the experiment using a
different clustering method, and found that instead of noticing the
number of ellipses, when that variable was removed participants instead
cued in on the differences in size and shape of the ellipses formed
using k-means clustering after the data generating procedure. That is,
participants could still detect the artificial nature of the induced
clusters using other features. While it can be difficult to get the data
generating method right, it is essential to conducting visual
experiments that generalize well beyond the effects shown in a single
data set or phenomenon. The time and effort invested in this step at the
outset of the experiment pays dividends when it allows for clear
generalization of the experimental results to an entire statistical
concept rather than a single data set.

@@

@@

@@

@@

It may be useful to ask a few more demographic questions about STEM
education level for studies which ask more of participants from a
mathematical standpoint; while lineup studies have not found strong
associations with those variables, lineup studies also do not require
participants to engage with the data presented in a chart in a way that
requires higher-order mathematical reasoning. This has allowed us to
make the argument to IRB that our research is exempt, as we do not
collect enough demographic information to identify participants,
however, this has occasionally come at some cost. Our recent study
examining the use of log scales and exponential data was conducted using
Prolific, which recruits participants from around the world; we required
only that participants were fluent in English to participate. It was
only after the experiment was completed that we realized that different
countries introduce logarithms as a concept at different points during
primary and secondary education; it might be that individuals in some
countries have much more experience with log scales than those educated
in the United States. We mention this only to point out that while every
experiment contains a few missed opportunities, it is worth giving
careful thought to the demographic questions asked of participants and
what information may be helpful during the analysis stage.

@@

\includegraphics{index_files/figure-pdf/fig-density-rug-1.pdf}

@@
